Computing solution space properties of combinatorial optimization problems via generic tensor networks

### 2022-05-23 paper/paper.tex
add ref
fix citations

\section{Tensor networks}\label{sec:tn}
A tensor network~\cite{Cirac2020, Orus2014} is composed of a collection of tensors, a collection of labels associated with tensor dimensions, and an indicator of which tensor is the output.
The operation to evaluate a tensor network is called \textit{contraction}, which is defined as a summation of tensor element products over the labels not appearing in the output tensor.
For example, the matrix multiplication is a special tensor network that can be represented as $C_{ik} = A_{ij}B_{jk}$, where $A, B$ and $C$ are matrices (two-dimensional tensors), $ik,ij$ and $ik$ are labels associated with them, and ``$=$'' is the indicator of $C$ being the output, while its contraction is defined as $C_{ik} = \sum_j A_{ij}B_{jk}$.
%can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
%Generalized Einstein's notation is often used to represent a tensor network, where we use a label in the subscript to represent a degree of freedom.
%One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
% where we use a label in the subscript to represent a degree of freedom.
% One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice, either both in input tensors (which will be summed over) or one in a input tensor and another in the output tensor.
% Hence a tensor network can be represented as a open simple graph,
% where a tensor is mapped to a vertex, a label shared by two input tensors is mapped to a normal edge and a label appears in the output tensor is mapped to an open edge.
% In this work, we adopt a more general form of tensor networks that do not restrict how many times a label can appear in tensors. 
The graphical representation of a tensor network is an open (meaning an edge can connect to external vertices) hypergraph, where an input tensor is mapped to a vertex and a label is mapped to a hyperedge that can connect an arbitrary number of vertices, while the labels appearing in the output tensor are open.
Our notation is a minor generalization of the standard tensor network notation used in physics as we do not restrict the number of times a label can appear in the tensors. 
While this generalized form is equivalent in representation power, it can have smaller contraction complexity as will be illustrated in \App{sec:tensorbad}.
% because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
%However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \App{sec:tensorbad}.

@@

Looking ahead, it could be possible to generalize the idea of generic programming to other algorithms that have certain algebraic structures such as those using the inclusion-exclusion principle or subset convolution~\cite{Fomin2013} and explore what new properties can be computed.
To this end, dynamic programming~\cite{Courcelle1990, Fomin2013} approaches could be considered.
Dynamic programming is closely related to a tropical tensor network~\cite{Liu2021}; for example, the Viterbi algorithm for finding the most probable configuration in a hidden Markov model can be interpreted as a matrix product state featured with tropical algebra, and the tropical tensor network in the main text is potentially equivalent to dynamic programming in finding an optimum solution.
Since dynamic programming has much broader applications, it would be interesting to extend the ideas from this paper to provide an algebraic interpretation for dynamic programming so that it can be used to compute other solution space properties beyond just finding an optimum solution.
%Moreover, since the independence polynomial is closely related to the matching polynomial~\cite{Levit2005}, the clique polynomial~\cite{Hoede1994}, and the vertex cover polynomial~\cite{Akbari2013},
%our algorithm to compute the independence polynomial can also be used to compute these graph polynomials.
%By adapting to the style of generic programming, we can implement all the algorithms without much effort.

@@

\section*{Acknowledgments}
We would like to thank Pan Zhang for sharing his python code for optimizing contraction orders of a tensor network.
We acknowledge Sepehr Ebadi and Leo Zhou for coming up with many interesting questions about independent sets and their questions strongly motivated the development of this project.
We thank Benjamin Schiffer for providing helpful feedback on the writing of this manuscript.
We thank Chris Elord for helping us write the fastest matrix multiplication library for tropical numbers, TropicalGEMM.jl. 
We would also like to thank a number of open-source software developers, including Roger Luo, Time Besard, Edward Scheinerman, and Katharine Hyatt
for actively maintaining their packages and resolving related issues voluntarily.
We acknowledge financial support from the DARPA ONISQ program (grant no.\ W911NF2010021), the Center for Ultracold Atoms, the National Science Foundation, the Vannevar Bush Faculty Fellowship, the U.S. Department of Energy (DE-SC0021013 and DOE Quantum Systems Accelerator Center (contract no.\ 7568717), 
%the Office of Naval Research, 
the Army Research Office MURI. We acknowledge the computation credits provided by Amazon Web Services for running the benchmarks and case studies. Jinguo Liu acknowledges funding support provided by QuEra Computing Inc.\ through a sponsored research program.

>>>>

\section{Tensor networks}\label{sec:tn}
A tensor network~\cite{Cirac2021, Orus2014} is composed of a collection of tensors, a collection of labels associated with tensor dimensions, and an indicator of which tensor is the output.
The operation to evaluate a tensor network is called \textit{contraction}, which is defined as a summation of tensor element products over the labels not appearing in the output tensor.
For example, the matrix multiplication is a special tensor network that can be represented as $C_{ik} = A_{ij}B_{jk}$, where $A, B$ and $C$ are matrices (two-dimensional tensors), $ik,ij$ and $ik$ are labels associated with them, and ``$=$'' is the indicator of $C$ being the output, while its contraction is defined as $C_{ik} = \sum_j A_{ij}B_{jk}$.
%can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
%Generalized Einstein's notation is often used to represent a tensor network, where we use a label in the subscript to represent a degree of freedom.
%One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
% where we use a label in the subscript to represent a degree of freedom.
% One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice, either both in input tensors (which will be summed over) or one in a input tensor and another in the output tensor.
% Hence a tensor network can be represented as a open simple graph,
% where a tensor is mapped to a vertex, a label shared by two input tensors is mapped to a normal edge and a label appears in the output tensor is mapped to an open edge.
% In this work, we adopt a more general form of tensor networks that do not restrict how many times a label can appear in tensors. 
The graphical representation of a tensor network is an open (meaning an edge can connect to external vertices) hypergraph, where an input tensor is mapped to a vertex and a label is mapped to a hyperedge that can connect an arbitrary number of vertices, while the labels appearing in the output tensor are open.
Our notation is a minor generalization of the standard tensor network notation used in physics as we do not restrict the number of times a label can appear in the tensors. 
While this generalized form is equivalent in representation power, it can have smaller contraction complexity as will be illustrated in \App{sec:tensorbad}.
% because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
%However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \App{sec:tensorbad}.

@@

Looking ahead, it could be possible to generalize the idea of generic programming to other algorithms that have certain algebraic structures such as those using the inclusion-exclusion principle or subset convolution~\cite{Fomin2013} and explore what new properties can be computed.
To this end, dynamic programming~\cite{Courcelle1990, Fomin2013} approaches could be considered.
Dynamic programming is closely related to a tropical tensor network~\cite{Liu2021}; for example, the Viterbi algorithm for finding the most probable configuration in a hidden Markov model can be interpreted as a matrix product state featured with tropical algebra, and the tropical tensor network in the main text is potentially equivalent to dynamic programming in finding an optimum solution.
Since dynamic programming has much broader applications, it would be interesting to extend the ideas from this paper to provide an algebraic interpretation for dynamic programming so that it can be used to compute other solution space properties beyond just finding an optimum solution.
It is also possible to extend this idea to other algebras. For example, generic semiring algebra has been used in computational linguishtics to compute interesting quantities of a given grammar and string.~\cite{Goodman1999}.
%Moreover, since the independence polynomial is closely related to the matching polynomial~\cite{Levit2005}, the clique polynomial~\cite{Hoede1994}, and the vertex cover polynomial~\cite{Akbari2013},
%our algorithm to compute the independence polynomial can also be used to compute these graph polynomials.
%By adapting to the style of generic programming, we can implement all the algorithms without much effort.

@@

\section*{Acknowledgments}
We would like to thank Pan Zhang for sharing his python code for optimizing contraction orders of a tensor network.
We acknowledge Sepehr Ebadi and Leo Zhou for coming up with many interesting questions about independent sets and their questions strongly motivated the development of this project.
We thank Benjamin Schiffer for providing helpful feedback on the writing of this manuscript.
We thank Chris Elord for helping us write the fastest matrix multiplication library for tropical numbers, TropicalGEMM.jl. 
We thank Jacob Miller for helpful discussions. 
We would also like to thank a number of open-source software developers, including Roger Luo, Time Besard, Edward Scheinerman, and Katharine Hyatt
for actively maintaining their packages and resolving related issues voluntarily.
We acknowledge financial support from the DARPA ONISQ program (grant no.\ W911NF2010021), the Center for Ultracold Atoms, the National Science Foundation, the Vannevar Bush Faculty Fellowship, the U.S. Department of Energy (DE-SC0021013 and DOE Quantum Systems Accelerator Center (contract no.\ 7568717), 
%the Office of Naval Research, 
the Army Research Office MURI. We acknowledge the computation credits provided by Amazon Web Services for running the benchmarks and case studies. Jinguo Liu acknowledges funding support provided by QuEra Computing Inc.\ through a sponsored research program.
